Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-31 paper.tex
Pass through discussion

\section{Discussion}
% Zarr is great
VCF is a central element of modern genomics, facilitating
the exchange of data in a large ecosystem of interoperating tools. 
Its current row-oriented form, however,
is fundamentally inefficient,
profoundly limiting the scalability of the current generation
of bioinformatics tools. Large scale VCF data cannot 
currently be 
processed without incurring a substantial economic
and likely environmental cost~\cite{grealey2022carbon}.
We have shown here that this is not a necessary situation,
and that greatly improved efficiency can be achieved by 
using more appropriate storage representations tuned
to the realities of modern computing. We have argued that 
Zarr provides a powerful basis for cloud-based
storage and analysis of large-scale genetic variation data.
We propose the VCF Zarr specification which losslessly
maps VCF data to Zarr, and provide an efficient and scalable
tool to perform conversion.

% We're not pretending is a solution to all problems and
% up front about weaknesses of the model.
Zarr provides pragmatic solutions to some of the more pressing 
problems facing the analysis of large-scale genetic variation
data, but it is not a panacea. Firstly, 
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the 
dimensions of fields are determined by their \emph{maximum}
dimension among all variants. In particular this is problematic
for fields like PL in which the dimension depends 
quadratically on the number of alleles (although practical
solutions have been suggested that we plan to 
implement~\cite{poterba2024scalable}).
Secondly, the design of 
VCF Zarr emphasises efficiency of analysis for a fixed 
dataset, and does not consider how samples (and the 
corresponding novel variants) should be added.
Thirdly, Zarr is not suitable for handling 
the unstructured data often included in VCF INFO 
fields, and would likely be a poor choice
for representing variant annotations, for example.

Nonetheless, there are numerous datasets that exist today
that would likely reap significant benefits from being deployed 
in a cloud-native fashion using Zarr. Object 
stores typically allow for individual objects (chunks, in
Zarr) to be associated with ``tags'', which can then be 
used to associate storage class, user access control
and encryption keys. 
Aside from the performance benefits
we have focused on here provided by Zarr, the ability
to (for example) use high-performance storage for commonly
used data such as the variant position and 
more cost-effective storage classes
for infrequently used bulk QC data should provide 
significant operational benefits.
Granular access controls would similarly allow non-identifiable
variant-level data to be shared relatively freely,
with genotype and other data more tightly controlled
as required.
Even finer granularity is possible if samples are grouped according
to access level within chunks (i.e, by padding partially filled 
chunks and using an appropriate sample mask). 
Delegating access control to the cloud provider in this way
and providing client applications direct access to
the data in an efficiently analysable form, 
largely negates the 
need for web API access~\cite{kelleher2019htsget} 
and cryptographic container formats~\citep{senf2021crypt4gh}
in this setting.

The draft VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
provided here are a necessary starting point for such cloud-native
biobank repositories and open up many possibilities,
but significant investment and development would be needed 
to provide a viable alternative to standard bioinformatics workflows. 
Two initial directions for development, however, may provide 
a sufficient starting point to both
greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England
and motivate further development.
The first direction is to provide compatibility with existing 
workflows via a ``\texttt{vcztools}'' command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) on a VCF Zarr dataset.
Such a tool would speed up some common queries by orders 
of magnitude, and reduce the need for user orchestration of 
operations among manually split VCF chunks (large VCF datasets
are typically split into hundreds of files; see the Genomics
England case study). Datasets could then be hosted in cloud
object stores, while still presenting file-like semantics
for existing workflows. This could provide an evolutionary path,
allowing established analysis workflows to coexist with new Zarr-native
approaches, working from the same primary data.

The second direction is then to develop new methods 
working directly with the Zarr data, and can leverage 
a rich ecosystem of powerful and efficient tools in 
multiple programming languages (see Methods).
The Python data science ecosystem, in particular,
has a rich suite of powerful tools~\cite[e.g.][]{
harris2020array,
mckinney2010data,
kluyver2016jupyter,
virtanen2020scipy} 
and multiple Zarr implementations (see Methods).
Of particular interest for large-scale datasets are
Dask~\cite{rocklin2015dask} and 
Xarray~\citep{hoyer2017xarray} which provide powerful abstractions
for distributed computing and can be
straightforwardly used to perform interactive analysis
of VCF Zarr datasets,
with computations transparently distributed over a cluster.
As shown in our benchmarks, lower-level computations
in which numerical operations are applied on a chunk-by-chunk
basis are also straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers can 
choose the appropriate level of abstraction and tool chain
on a case-by-case basis.

[NOTE: still trying to figure out how to string this together. 
Want to finish on a positive, not too technical note.]

and is gaining increasing
popularity in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Zarr has the potential to act as a unified, cloud-native 
efficient storage layer, freeing the next generation of 
bioinformatics tools
from the traditional burdens of text parsing and format
conversion.

>>>>

\section{Discussion}
% Zarr is great
VCF is a central element of modern genomics, facilitating
the exchange of data in a large ecosystem of interoperating tools. 
Its current row-oriented form, however,
is fundamentally inefficient,
profoundly limiting the scalability of the present generation
of bioinformatics tools. Large scale VCF data cannot 
currently be 
processed without incurring a substantial economic
(and environmental~\cite{grealey2022carbon}) cost.
We have shown here that this is not a necessary situation,
and that greatly improved efficiency can be achieved by 
using more appropriate storage representations tuned
to the realities of modern computing. We have argued that 
Zarr provides a powerful basis for cloud-based
storage and analysis of large-scale genetic variation data.
We propose the VCF Zarr specification which losslessly
maps VCF data to Zarr, and provide an efficient and scalable
tool to perform conversion.

% We're not pretending is a solution to all problems and
% up front about weaknesses of the model.
Zarr provides pragmatic solutions to some of the more pressing 
problems facing the analysis of large-scale genetic variation
data, but it is not a panacea. Firstly, 
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the 
dimensions of fields are determined by their \emph{maximum}
dimension among all variants. In particular this is problematic
for fields like PL in which the dimension depends 
quadratically on the number of alleles (although practical
solutions have been suggested that we plan to 
implement~\cite{poterba2024scalable}).
Secondly, the design of 
VCF Zarr emphasises efficiency of analysis for a fixed 
dataset, and does not consider how samples (and the 
corresponding novel variants) should be added.
Thirdly, Zarr works best for numerical data of a fixed 
dimension, and therefore may not suitable for representing
the unstructured data often included in VCF INFO fields.

Nonetheless, there are numerous datasets that exist today
that would likely reap significant benefits from being deployed 
in a cloud-native fashion using Zarr. Object 
stores typically allow for individual objects (chunks, in
Zarr) to be associated with ``tags'', which can then be 
used to associate storage class, user access control
and encryption keys. 
Aside from the performance benefits
we have focused on here provided by Zarr, the ability
to (for example) use high-performance storage for commonly
used data such as the variant position and 
more cost-effective storage classes
for infrequently used bulk QC data should provide 
significant operational benefits.
Granular access controls would similarly allow non-identifiable
variant-level data to be shared relatively freely,
with genotype and other data more tightly controlled
as required.
Even finer granularity is possible if samples are grouped by
access level within chunks (padding partially filled 
chunks as needed and using an appropriate sample mask). 
Providing client applications direct access to 
the data over HTTP
and delegating access control to the cloud provider
makes custom web APIs~\cite{kelleher2019htsget} 
and cryptographic container formats~\citep{senf2021crypt4gh}
largely unnecessary.

The VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
provided here are a necessary starting point for such cloud-native
biobank repositories and open up many possibilities,
but significant investment and development would be needed 
to provide a viable alternative to standard bioinformatics workflows. 
Two initial directions for development, however, may quickly
yield sufficient results to both greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England
and motivate further research and development.
The first direction is to provide compatibility with existing 
workflows via a ``\texttt{vcztools}'' command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) on a VCF Zarr dataset.
Such a tool would speed up some common queries by orders 
of magnitude, and reduce the need for user orchestration of 
operations among manually split VCF chunks (large VCF datasets
are typically split into hundreds of files; see the Genomics
England case study). Datasets could then be hosted in cloud
object stores, while still presenting file-like semantics
for existing workflows. This could provide an evolutionary path,
allowing established analysis workflows to coexist with new Zarr-native
approaches, working from the same primary data.

The second natural direction for development is to create
these Zarr-native applications, which can take advantage
of the efficient data representation
across multiple programming languages (see Methods).
The Python data science ecosystem, in particular,
has a rich suite of powerful tools~\cite[e.g.][]{
mckinney2010data,
lam2015numba,
kluyver2016jupyter,
harris2020array,
virtanen2020scipy} 
and is increasingly
popular in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Dask~\cite{rocklin2015dask} and 
Xarray~\citep{hoyer2017xarray} provide powerful abstractions
for large-scale data analysis and can be
used to perform interactive analysis
of large VCF Zarr datasets,
with computations transparently distributed over a cluster.
As shown in our benchmarks, lower-level computations
in which optimised kernels are applied on a chunk-by-chunk
basis are also straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers can 
choose the appropriate level of abstraction and tool chain
on a case-by-case basis.

Zarr has the potential to act as a unified, cloud-native 
efficient storage layer, freeing the next generation of 
bioinformatics tools
from the traditional burdens of text parsing and format
conversion.
